## Reproducible Research: Peer Assessment 1

### Loading and preprocessing the data
Load Libraries
```{r}
library(ggplot2)
library(xtable)
library(plyr)
library(scales)
```

To load the data, I unzip the activty folder and read in the actvitiy file. 
I also preprocessed the data by changing the class of the date and interval variables.

```{r Read_Activity}
unzip("activity.zip")
activity <- read.csv("activity.csv")
activity$date <- as.Date(activity$date,"%Y-%m-%d")
activity$interval <- as.POSIXct(sprintf("%04d",activity$interval), format= "%H%M")
activity$interval <- format(activity$interval, format="%H:%M")
```
### What is mean total number of steps taken per day?

#### 1. Make a histogram of the total number of steps taken each day

The count of steps per day is calculated to create a histogram of the total 
number of steps taken each day shown below.
```{r count steps per day} 
ActSum <- ddply(activity, .(date), summarise, StepTot = sum(steps, na.rm=T))
```
```{r Histgram_Steps_per_Day, fig.height=5, fig.width=8}
ggplot(ActSum, aes(date, StepTot)) + 
	labs(x= "Date", y="Steps per Day") +
	geom_bar(stat="identity", fill="steelblue4") +
	ggtitle("Histogram of Total # of Steps per Day")
```	

#### 2. Calculate and report the mean and median total number of steps taken per day

```{r mean_medianStepsPerDay, results='asis'} 
ActCenter <- ddply(activity, .(date), summarise, 
					Step_Mean = round(mean(steps, na.rm=T),4),
					Step_Median = median(steps, na.rm=T))
ActCenter$date <- as.character(ActCenter$date)
print(xtable(ActCenter), type="html")
```

### What is the average daily activity pattern?

#### 1. Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days

```{r Avg steps per interval} 
ActMeanInterval <- ddply(activity, .(interval), summarise, 
							StepMean= mean(steps, na.rm=T))
ActMeanInterval$interval <- as.POSIXct(sprintf("%04s",as.character(ActMeanInterval$interval)), 
								format= "%H:%M")
```
```{r Time Series-Interval by Avg Step, fig.height=5, fig.width=7}
ggplot(ActMeanInterval, aes(interval, StepMean)) + 
	labs(x= "5-Minute Interval", y="Average Steps per Interval") +
	geom_line() + 
	scale_x_datetime(breaks = date_breaks("2 hours"),
		labels = date_format("%H:%M")) +
	ggtitle("Time Series plot: 5-minute Interval by Average Steps")
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxInterval <- ActMeanInterval[ActMeanInterval$StepMean %in% max(ActMeanInterval$StepMean,na.rm=T),]
```
The 5-minute interval, `r format(maxInterval$interval, "%H:%M")`, has the maximum average number of steps.

### Imputing missing values

#### 1. Calculate and report the total number of missing values in the dataset

``` {r}
missing <- sum(is.na(activity$steps))
```
The total number of missing values in the activity dataset is `r missing`.

#### 2. Devise a strategy for filling in all of the missing values in the dataset. 

To fill in the missing values I will use the average of the 5-minute interval. 
I will use the ActMeanInterval dataset since the average 5-minute interval (over all days) 
was already calculated.

#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
ActMeanInterval$interval <- format(ActMeanInterval$interval, format="%H:%M")
Actmerge <- merge(activity, ActMeanInterval, by.x="interval", by.y="interval") 
Actmerge$steps <- ifelse(is.na(Actmerge$steps), Actmerge$StepMean, Actmerge$steps)
```

#### 4. Make a histogram of the total number of steps taken each day 

```{r} 
ActSum2 <- ddply(Actmerge, .(date), summarise, StepTot = sum(steps))
```
```{r Histgram_Steps_per_Day_Imputed, fig.height=5, fig.width=8}
ggplot(ActSum2, aes(date, StepTot)) + 
	labs(x= "Date", y="Steps per Day") +
	geom_bar(stat="identity", fill="violetred4") +
	ggtitle("Histogram of Total # of Steps per Day (Missing Values Imputed)")
```	
#### Calculate and report the mean and median total number of steps taken per day. 

```{r, results='asis'} 
ActCenter2 <- ddply(Actmerge, .(date), summarise, 
					Step_MeanImputed = round(mean(steps, na.rm=T),4),
					Step_MedianImputed = median(steps, na.rm=T))
ActCenter2$date <- as.character(ActCenter2$date)
print(xtable(ActCenter2), type="html")
```
#### Do these values differ from the estimates from the first part of the assignment? 
Yes, the mean and median  values differ from the estimates in first part of the assignment

#### What is the impact of imputing missing data on the estimates of the total daily number of steps?
The 8 days with missing daily steps have the same estimates for total daily steps in the imputed dataset.

### Are there differences in activity patterns between weekdays and weekends?

#### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
Actmerge$Weekdays <- ifelse(weekdays(Actmerge$date) %in% c("Saturday","Sunday"), "weekend", "weekday")
```
#### 2. Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days. 
```{r} 
ActMeanInt_imp <- ddply(Actmerge, c(.(interval), .(Weekdays)), summarise, 
						StepMeanImp= mean(steps, na.rm=T))
ActMeanInt_imp$interval <- as.POSIXct(sprintf("%04s",as.character(ActMeanInt_imp$interval)), 
								format= "%H:%M")
```

```{r Time Series Comparison-Weekdays vs Weekend, fig.height=5, fig.width=7}
ggplot(ActMeanInt_imp, aes(interval, StepMeanImp)) + 
	labs(x= "5-Minute Interval", y="Average Steps per Interval") +
	geom_line() + facet_wrap(~Weekdays, nrow=2) +
	scale_x_datetime(breaks = date_breaks("2 hours"),
		labels = date_format("%H:%M")) +
	ggtitle("Time Series plot Comparison: 5-minute Interval by Average Steps")
```